---
title: "Quiz 1 Fall 2022"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Investigating insect declines

Recent studies have noted large declines in insect abundances, but these studies have mainly
focused on evidence from Europe.  Crossley at al (2021) investigated this issue using 
publicly-available data from Long-term Ecological Research (LTER) projects.

Data can be loaded using 

```{r}
time_series <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/data_from_crossley_et_al_2021.csv",
                       stringsAsFactors = T)
```

Create a properly-labeled plot that shows how change over time (measured as slope, we'll get 
to that later in class) compares among LTERs. Make sure you show variation somehow!
Note the column of interest is **b** and groups are defined in the **LTER** 
column. Multiple options exist.
