---
title: "6. Continuous tests 2 populations"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
    toc_depth: 6
---

Before doing this, review the **Extending continuous tests to 2 samples** lecture set slides from 
https://sites.google.com/view/biostats/lessons/2-sample-tests-for-continuous-data and
the  **6_2_sample_continuous tests.R** script in the lecture_files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats).


Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github

## Examples

### Paired tests and rearranging data

Paired tests assume there is a one-to-one relationship between measurements in each
group (before/after, twin 1/twin 2, etc).
Note for paired tests the arrangement of the data matters. For example, if paired
data is stored so that each row has information from a single pair, we need a
different test format than if each row only has information from a single unit. 
In general, when a row only has information from one unit, we call it long-formatted
data. Wide-formatted data has information from multiple units on each row.  This
will get more complicated when we measure multiple traits/things on each unit, 
but its a start.

#### t-test

Let's first load a long-formatted dataset.

```{r}
cholesterol <- read.table("http://www.statsci.org/data/general/cholestg.txt", header = T,
                          stringsAsFactors = T)
head(cholesterol)
str(cholesterol)
```

Note this is long format data.  Day is also an integer here. For our purposes, lets consider it a factor (near/far
from heart attack).

```{r}
cholesterol$day <- as.factor(cholesterol$day)
library(plyr)
cholesterol$when <- revalue(cholesterol$day, c("2"="soon_after", "4" = "few_days",
                                               "14" = "weeks_later"))
```

To run a t-test on this format, we would use the formula notation and only focus 
on 2 periods (since its a t-test!). Note the *%in%* operator lets us search for 
a list of outcomes to match (instead of *==*, which matches one).

```{r, error=T}
t.test(cholest ~ when, cholesterol[cholesterol$when %in% c("soon_after", 
                                                           "weeks_later"),],
       paired = T)
```
This does not work (note I had to add an R chunk option, *error = T *, to make it
knit!).  Why?  The error mentions an issues with length.  Look at each piece

```{r}
 cholesterol[cholesterol$when %in% c("soon_after"), "cholest"]
 cholesterol[cholesterol$when %in% c("weeks_later"), "cholest"]
```

Note we have many NAs. Further inspection (look at the whole cholesterol dataset
by clicking on it in environment) shows we have 2 groups, which isn't helping. 
R often exclude's NAs by default. Try

```{r}
options("na.action")
```

It likely says "na.omit".  For paired tests in formula format this is an issue because
it looks for matched pairs and *then* excludes NAs. To make this work, let's 
make sure to send the function everything using the added na.action argument set to
"na.pass". 

```{r}
t.test(cholest~day, cholesterol[cholesterol$when %in% c("soon_after","weeks_later"),],
       paired = T, na.action = "na.pass")
```

Note this is same as 

```{r}
t.test(cholesterol[cholesterol$day == "2", "cholest"], 
       cholesterol[cholesterol$day == "14", "cholest"],
       paired = T)
```

which does not have the *NA* issue.  This setup is better for wide format data.

#### long to wide 

Let's make the cholesterol dataset wide usin the dcast function from the *reshape2* 
package. The formula notes row ~ columns.  

```{r}
library(reshape2)
#formula gives row ~ columns
#get daily cholesterol for every patient
cholesterol_wide <- dcast(data = cholesterol, formula = patient ~ day, 
                          value.var ="cholest" )
head(cholesterol_wide)
```

We could evaluate this using

```{r, error = T}
t.test(cholesterol_wide$2, cholesterol_wide$14, paired = T)
```
Another error!  R doesn't like columns to be named with numbers. Let's fix this.

```{r}
names(cholesterol_wide)[2:5] <- paste ("day", names(cholesterol_wide)[2:5], 
                                       sep="_")
head(cholesterol_wide)
```

Now try

```{r}
t.test(cholesterol_wide$day_2, cholesterol_wide$day_14, paired = T)
```
#### Wilcoxon test

Compare these to Wilcoxon test

```{r}
wilcox.test(cholest~day, cholesterol[cholesterol$when %in% c("soon_after","weeks_later"),],
       paired = T, na.action = "na.pass")
wilcox.test(cholesterol_wide$day_2, cholesterol_wide$day_14, paired = T)
```

#### Sign test

or Sign tests (which only take wide format and only work with paired data!)

```{r}
library(BSDA)
SIGN.test(cholesterol_wide$day_2, cholesterol_wide$day_14)
```

### Not paired

What if we had measured different people a few days before and several weeks after
a heart attack?  Then the data is not paired! Examples for all tests shown here.

#### t-tests

Note by default we get a Welsh test, which assumes variance is not the same among
the groups.

```{r}
t.test(cholest~day, cholesterol[cholesterol$when %in% c("soon_after","weeks_later"),])
t.test(cholesterol_wide$day_2, cholesterol_wide$day_14)
```
#### Wilcoxon test

```{r}
wilcox.test(cholest~day, cholesterol[cholesterol$when %in% c("soon_after","weeks_later"),])
wilcox.test(cholesterol_wide$day_2, cholesterol_wide$day_14)
```
#### Bootstrap

We also have options that don't rely on repeated sampling.  My bootstrap function
only works for wide format data (unless you manually subtract outcomes... and omit
NAs...)

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/bootstrapjsg.R")
bootstrapjsg(cholesterol_wide$day_2, cholesterol_wide$day_14)

cholesterol_difference <- na.omit(cholesterol_wide$day_2 - cholesterol_wide$day_14)
bootstrapjsg(cholesterol_difference)
```
Note again how pairing helps!

#### Permutation

Note the permutation tests only works on unpaired data.

```{r}
library(coin)
independence_test(cholest~day, 
                  cholesterol[cholesterol$when %in% c("soon_after","weeks_later"),])
```
### More on rearranging data

If you need to re-arrange data **AND** summarize it in same step (not one-to-one
match when you rearrange), you'll need to tell R how to summarize it. For example,
we could find the average cholesterol for each patient (note there are 2 groups that
label each patient 1 ). Note I made a function to remove NAs, which otherwise cause
issues.

```{r}
#use fun.aggregate to get other option
#get average cholesterol per patient
#simple function to exclude na's (other option is to melt first and then drop them)
meannona <- function (x) mean(x, na.rm=T)
cholesterol_wide_summary <- dcast(data = cholesterol, formula = patient ~ ., #. means no variable, ... means all variables
                          value.var ="cholest", fun.aggregate = meannona )
head(cholesterol_wide_summary)
```

You can also name output column by putting in quotes in formula

```{r}
cholesterol_wide_summary <- dcast(data = cholesterol, formula = patient ~ "average_cholest", #. means no variable, ... means all variables
                          value.var ="cholest", fun.aggregate = meannona )
head(cholesterol_wide_summary)
```


#### Wide to long

If you have data in wide format

```{r}
cholesterol <- read.table("http://www.statsci.org/data/general/cholestr.txt", header = T)
head(cholesterol)
```

you can instead use the *melt* function (also in the *reshape2* package).  The
id.vars argument lists independent values to keep. The 
measure.vars variable is what you are measuring (not used here, and used rarely). 
The variable.name argument lists what to label things you group, and the value.name 
arguments gives name to value output

```{r}
cholesterol_long <- melt(cholesterol, id.vars =c())
head(cholesterol_long)

#name outcomes
cholesterol_long <- melt(cholesterol, id.vars =c(), variable.name = "day", 
                         value.name = "cholesterol")
head(cholesterol_long)
```

If you have more variables, you can melt them all to a single column (most useful
when faceting plots!).

```{r}
sport <- read.table("http://www.statsci.org/data/oz/ais.txt", header = T)
sport_melted <- melt(sport, id.vars = c("Sex", "Sport"),
                     variable.name = "measure", 
                     value.name = "value")
head(sport_melted)
```

## Practice

### For the following problems, use the data to conduct each noted test (note some sample sizes may be too small for these to all be good ideas!)


#### 1

1. The following data are human blood clotting times (in minutes) of individuals
given one of two different drugs. 

|Drug B | Drug G
|:-: |:-:
|8.8 | 9.9
|8.4 | 9.0
|7.9 | 11.1
|8.7 | 9.6
|9.1 | 8.7
|9.6 | 10.4 
|    | 9.5

Test the hypothesis that the mean clotting times
are equal for the two groups

* Estimating the variance from the data 
* Using rank transform analysis 
* Using a permutation test 
* Using a bootstrap test 

#### 2

2. Data on plant heights (in cm) for plants grown with a new and old formulation 
of fertilizer can be found at
https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv . 
Use the data to test the hypothesis that there is no difference in mean plant
heights for the two groups 

* Estimating the variance from the data 
* Using rank transform analysis 
* Using a permutation test 
* Using a bootstrap test 

Note you may need to cast this dataframe into a dataframe with a column for 
old and new outcomes.  You’ll have to add a unique identifier. Try something like

```{r, eval=FALSE}
fertilizer$ID <- 1:nrow(fertilizer)
fertilizer_wide <- dcast(fertilizer, ID~fertilizer, value.var = "height")
```

#### 3

3. Metabolic rates were measured in animals before and after administering a drug. Data is
available at 

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/metabolic_rates.csv

Test the hypothesis that the drug has no effect on metabolic rate with using 

* a Student's t test
* a rank test
* a binary test
* bootstrapping

#### 4

4. Endurance of the above animals was also measured. Data is available at

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/endurance.csv

Test the hypothesis that the drug has no effect on using 
* a Student's t test
* a rank test
* a binary test
* bootstrapping

### For the following problems, make sure you can justify the test you choose to use and carry it out

#### 5

5. Using data from Australian athletes (http://www.statsci.org/data/oz/ais.html 
for details), determine if the average male training at the Australian Institute 
of Sport differs in from the average female training at the site in mercury
concentration (Hg). 

#### 6

6. Laureysens et al. (2004) measured metal content in the wood of 13 poplar 
clones growing in a polluted area, once in August and once in November.
Concentrations of aluminum (in micrograms of Al per gram of wood) are shown 
below. (From Handbook of Biological Statistics). Has metal content changed over
time?

| Clone | August | November
|:-:| :-:|:-:
|Columbia River | 18.3 |12.7
|Fritiz Pauley | 13.3 |11.1
|Hazendas | 16.5 |15.3
|Primo | 12.6 | 12.7
|Raspalje | 9.5 |10.5
|Hoogvorst |13.6 |15.6
|Balsam Spire | 8.1 |11.2
|Gibecq   |8.9 |14.2
|Beaupre | 10.0 |16.3
|Unal |8.3 |15.5
|Trichobel |7.9 | 19.9
|Gaver | 8.1 | 20.4
|Wolterson | 13.4 |36.8

#### 7

7. Data was collected on cholesterol levels following heart attacks.  Data
available at 
http://www.statsci.org/data/general/cholestg.txt 

Note to read a .txt file you use,
read.table("http://www.statsci.org/data/general/cholestg.txt", header = T, 
stringsAsFactors = T)
Check the defaults to understand the argument.  
More info on the data is @
http://www.statsci.org/data/general/cholest.html
Is there evidence cholesterol levels changed between day 2 and day 4?  
* **THINK ABOUT RELABELLING DAY AS A FACTOR (WHY?)?**

